---
title: "MsQualityUtils: Creation of MsExperiment objects for MsQuality-based QC metric calculation"
author:
 - name: Thomas Naake
   mail: thomas.naake@embl.de, thomasnaake@googlemail.com
   affiliation: European Molecular Biology Laboratory, Meyerhofstrasse 1, 69117 Heidelberg
package: MsQualityUtils
      
output:  
    BiocStyle::html_document:
        toc_float: true
bibliography: MsQualityUtils-citations.bib
vignette: >
    %\VignetteIndexEntry{QC for metabolomics and proteomics data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteKeywords{Mass Spectrometry, MS, Metabolomics, Proteomics, Visualization, QC}
    %\VignettePackage{MsQualityUtils-vignette}
    %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, include=FALSE, echo=FALSE, cache=FALSE}
library("knitr")
opts_chunk$set(stop_on_error = 1L)
suppressPackageStartupMessages(library("MsQuality"))
suppressPackageStartupMessages(library("MsQualityUtils"))
suppressPackageStartupMessages(library("Spectra"))
```

# Introduction {#sec-intro}

Data quality assessment is an integral part of preparatory data analysis 
to ensure sound biological information retrieval. 


We present here the `MsQualityUtils` package, which provides functionality to 
create `MsExperiment` objects from mzML and MetIDQ-derived files. 
The `MsExperiment` files can subsequently be used to calculate quality metrics 
at the per-sample level using the 
[`MsQuality`](https://github.com/tnaake/MsQuality) package. 

`MsQuality` relies on the [`mzQC`](https://github.com/HUPO-PSI/mzQC) 
framework of quality metrics defined by the Human Proteome 
Organization-Proteomics Standards Intitiative (HUPO-PSI). 

The `MsQuality` package is built upon the `Spectra` and the `MsExperiment` package.
Metrics will be calculated based on the information stored in a 
`Spectra` object, thus, the spectral data of each sample should be stored
in one `Spectra` object. The `MsExperiment` serves as a container to 
store the mass spectral data of multiple samples. `MsQuality` enables the user
to calculate quality metrics both on `Spectra` and `MsExperiment` objects. 
The `MsQualityUtils` package provides functionality to create `MsExperiment` 
objects that store the (spectral) information of mzML or MetIDQ files. 

In this vignette, we will 
(i) create an `MsExperiment` object from MetIDQ-derived mass spectrometry data  
and (ii) create an `MsExperiment` object from mzML mass spectrometry data.

# Installation

To install `MsQualityUtils`, you have first to install the 
[devtools](http://cran.r-project.org/web/packages/devtools/index.html) package: 

```r
install.packages("devtools")
library("devtools")
```

Install the `MsQualityUtils` package then via
```r
install_github("tnaake/MsQualityUtils")
```

This will install this package and all eventually missing dependencies.

# Questions and bugs {-}

`MsQualityUtils` is currently under active development. If you discover any 
bugs, typos or develop ideas of improving `MsQuality` feel free to raise an 
issue via [GitHub](https://github.com/tnaake/MsQualityUtils) 
or send a mail to the developer.

# Prepare the environment

Load the `Spectra`, the `MsExperiment` and the `MsQuality` package.

```{r load_MsExperiment_Spectra}
library("Spectra")
library("MsExperiment")
library("MsQuality")
```

# Create `MsExperiment` from MetIDQ files 

MetIDQ$^{\mathrm{TM}}$ xlsx files are created by the Biocrates 
MetIDQ$^{\mathrm{TM}}$. They store information on metabolite concentration
in their cells. The `MsQualityUtils` package provides functionality to 
retrieve the information stored in the MetIDQ$^{\mathrm{TM}}$ xlsx files
and create `Spectra` objects from this. Typically, some of the measured 
metabolites are separated by liquid chromatography. We can pass a
`SummarizedExperiment` object `rt` to the function `` that stores 
retention time values in the `assay` slot. If there is no object `rt` passed
to the function, an internal `rt` object will be used and retention times
are taken therefrom.

First, we will create a list of `Spectra` objects. The list has two entries:

 - the first entry contains all features found in the xlsx files (irrespective 
   of their retentiont time),
 - the second entry contains the features that are separated by a 
   chromatographic method, these features have information on retention time
   values.

The function will create `Spectra` objects from all MetIDQ files that are 
found in `path` (`createListOfSpectra` will combine all MetIDQ files found in
`path`). 

```{r sps_metidq}
path <- system.file("metidq", package = "MsQualityUtils")
sps_l <- createListOfSpectra(type = "MetIDQ", path = path, sheet = 1)
```

In a next step, we create a list of `MsExperiment` objects from the list of 
`Spectra` objects. The first entry of the list contains the `MsExperiment` 
object of features irrespective of their retention time. The second entry 
contains the `MsExperiment` object of features that are separated by a
chromatographic method, these features have information on retention 
time values. 

```{r msexp_metidq}
msexp_metidq_l <- createListOfMsExperimentFromListOfSpectra(sps_l = sps_l)
```


# Create `MsExperiment` from mzML files

Here, we load the example 
files from the `sciex` data set of the `msdata` package and create a list 
of one `Spectra` object from the two provided `mzML` files. The function
will create a `Spectra` object containing all `mzML` files that are 
found in the specified `path`.

```{r sps_mzml}
path <- system.file("sciex", package = "msdata")
sps_l <- createListOfSpectra(type = "mzML", path = path)
```

In a next step, we calculate a list of one `MsExperiment` object from the 
list of `Spectra` objects:
```{r msexp_mzml}
msexp_mzml_l <- createListOfMsExperimentFromListOfSpectra(sps_l = sps_l)
```


# Calculating the quality metrics on `MsExperiment` objects using `MsQuality`

The objects from the `MsQualityUtils` functions can directly be used 
by the `MsQuality` package.

The function `calculateMetrics` will iterate over the samples (referring
to rows in `sampleData(msexp))`) and calculate the quality metrics on the
corresponding `Spectra` objects.

```{r calculateMetrics}
library("MsQuality")
msexp_mzml <- msexp_mzml_l[[1]]
metrics_msexp <- calculateMetrics(object = msexp_mzml,
    metrics = qualityMetrics(msexp_mzml),
    relativeTo = "Q1", msLevel = 1L)
metrics_msexp
```

When passing an `MsExperiment` object to `calculateMetrics` a `data.frame`
object is returned with the samples (derived from the rownames of 
`sampleData(msexp)`) in the rows and the metrics in columns. 

We will show here the objects `metrics_rplc_msexp` and `metrics_hilic_msexp`
```{r paged_table_metrics, eval = TRUE, echo = FALSE}
print("metrics_msexp")
rmarkdown::paged_table(as.data.frame(metrics_msexp))
```

# Appendix {-}

## Session information {-}

All software and respective versions to build this vignette are listed here:

```{r session,eval=TRUE, echo=FALSE}
sessionInfo()
```

## References

